A generalization of the Bradley-Terry model for draws in chess

### 2020-09-15 inst/jebo_chess_rebuttal.txt

Referee Report

The referee's comments are preceded by a chevron (">") and my response
is included inline.  I have changed line-breaks in the interests of
clarity.

Overall Response

I am delighted that this insightful referee "gets" the manuscript.
The referee makes a number of observations and suggestions, all of
which I agree with, and I have made changes to the submission in
response, colour coded for ease of reference.  I have tried very hard
to preserve the brevity and pleasing structure that the referee
explicitly notes.  In short, the revised manuscript is a stronger and
more widely applicable document and I commend it to you.

@@

My view is that this damping feature is sometimes desirable (old games
are surely less informative than more recent ones) but uncontrolled in
the sense that the time-constant is not explicit or controllable, and
the Elo system uses an arbitrary points system with arbitrary cut-offs
[as opposed to the objective Bradley-Terry system].

> 
> Precise estimation will hinge on a large number of games, but this
> will be in tension with the assumption of time-invariant
> player-strengths.  The author should explore this tension.
>

I am in total agreement with this observation.  My thinking on this
issue has been strongly affected by my recent work applying
Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
start their careers as weak competitors, become stronger with
experience, and finally become uncompetitive due to age [although
frankly Lewis Hamilton at 35yo seems to buck this general trend, as
recent events in Tuscany would indicate].  I would suggest that chess
players have a similar career arc, although Formula 1 Teams typically
take a dim view of underperforming drivers---whom they rapidly and
unceremonially drop.  The career arc for these competitors would have
a truncated right hand side, unlike chess players who typically
decline gradually over decades [although I would observe that
Kasparov's current FIDE rating of 2812 is not that much less than his
peak of 2851].

Specifically, the dataset in question stretches from 1981 to 2017, a
very wide range.  Generalized Bradley-Terry, together with reified
entities, is capable of handling strengths varying over a career: one
would introduce a "mid-career monster", or perhaps an "experience
monster".  I would add here that monsters' strength is inherently
positive, so one could deal with waning strength with age by
introducing a "youth monster" who helps younger players.  But such
analysis is, I would suggest, firmly in the category of further work.
I have added a discussion of this issue to the manuscript (RED), in
two parts: firstly when introducing the dataset, and secondly under
Further Work.

### 2020-09-15 inst/jebo_chess_revision2.tex
rephrasing

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
  2017.  During this time, the players' strengths might have changed
  substantially.  Statistical assessment of time-varying strength is
  possible with the reified entity technique and this is discussed
  under further work below.  However, we may adopt one of two
  approaches: either we make the working assumption that the strengths
  do not change, or alternatively we wish to make inferences about the
  players' strength {\em averaged over a playing career}.  Either
  way,} Equation~\ref{likelihood_WD} becomes

>>>>

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
  2017.  During this time, the players' strengths might have changed
  substantially.  Statistical assessment of time-varying strength is
  possible with the reified entity technique and this is discussed
  under further work below.  However, we may treat the strengths as
  constant by adopting one of two approaches: either we assume that
  the strengths do not change, or alternatively we make inferences
  about the players' strength {\em averaged over a playing career}.
  Either way,} Equation~\ref{likelihood_WD} becomes
